Report on the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4)

### 2017-05-03 WSSSPE4_report/WSSSPE4_report.tex
Changes for section on papers and demos

\todo{group by type - idea, position, experience, demo}
\begin{itemize}
\item \textbf{Michael Heroux: \emph{Idea paper: Sustainable \& Productive:
Improving Incentives for Quality Software}}~\cite{Heroux:2016ws}.
The author presents that improving quality of scientific software is a measure
to address the increased complexity in both modeling and software. To support
quality, it is essential to provide Computational Science and Engineering teams
with publication venues, funding, and  professional recognition. Broad
improvements in sustainability and productivity can be only achieved when
publishers, funding agencies and employers raise their expectations for
software quality. Additionally, software community leaders are in the unique
position to positively impact software quality by working on establishing
incentives that will spur creative and novel approaches to improve developer
productivity  and software sustainability.

@@

\item \textbf{Olivier Philippe, Simon Hettrick, and Neil Chue Hong:
\emph{Experience Paper: Preliminary analysis of a survey of UK Research
Software Engineers}}~\cite{Philippe:2016ws}.
This paper presents results from a survey conducted on the new role of research
software engineers in academia. The survey provides demographic information
about the education, field, gender, job satisfaction, and career plans of the people
in this community. The outcome is that members of the community  were found to be
highly educated, derive mainly from the hard sciences, and are predominantly male.
Respondents reported satisfaction in their jobs, but indicated that career progression
is both difficult and opaque. The authors urge a continued discussion about the experience
of research software engineers and recommend further investigation into this
important community.

@@

\item \textbf{Anshu Dubey and Katherine Riley: \emph{Experience Paper: Software
Engineering and Community Codes Track in ATPESC}}~\cite{Dubey1:2016ws}.
The Argonne Training Program in Extreme Scale
Computing (ATPESC) was started by Argonne National Laboratory with the objective
of expanding the ranks of better prepared users of high-performance computing (HPC) machines.
One of the unique aspects of the program was inclusion of a
software engineering and community codes track. The inclusion
was motivated by the observation that the projects with a good
software process were better able to meet their scientific goals.
The paper presents the experience in running the software
track, the motivation, reception, and evolution of the track over the years.
The authors recommend the inclusion of similar tracks in other HPC-oriented training programs.

@@

\item \textbf{Anshu Dubey and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
of Software for Scientific Simulation}}~\cite{Dubey2:2016ws}.
The authors discuss the specifics of the lifecycle of scientific software.
While the software lifecycle is a well-researched topic, one class of projects,
research software, has received relatively
little attention from lifecycle researchers. This
paper formalizes a lifecycle model
for end-to-end scientific application software, featuring loose
coupling between submodels for development of infrastructure
and scientific capability. The authors also invite input from stakeholders
to converge on a model that captures the complexity of this
development processes and provides needed lifecycle guidance to
the scientific software community.

\item \textbf{Francisco Queiroz and Rejane Spitz: \emph{Position Paper:
Collaborative Gamification Design for Scientific Software}}~\cite{Queiroz:2016ws}.
The paper focuses on usability design and presents that gamification, a design
trend for improving scientific software usability, for scientific software should
be facilitated by an open, collaborative design process supported by
conversational media. This approach is compatible with
qualities often attributed to computational science community
regarding openness and collaboration between members of varied
professional backgrounds. They exemplify the use of conversational media for
collaborative design and expect the synergy between collaborators to result in
better usability, greater user acceptance, and adequacy to requirements,
obtaining optimal design solutions in a sustainable way.

\item \textbf{Bruce Childers, Jack Davidson, Wayne Graves, Bernard Rous, and
David Wilkinson: \emph{Position Paper: Active Curation of Artifacts is Changing
the Way Digital Libraries will Operate}}~\cite{Childers:2016ws}.
The paper presents the implications of ``active curation systems'' that provide
executable access to
artifacts and experiments behind published results and enables
their reuse. These systems allow changing and repeating experiments to understand
how an innovation behaves in conditions beyond the ones described in a paper and
address reproducibility challenges. Active curation systems
also  enable accountability and accelerate research progress  by
giving access to complete experimental details. As these systems
take  hold,  it  is  important  to  understand  their  capabilities  and
how digital libraries should be integrated with them. The authors describe a
study underway to explore how best
to do this integration.

\item \textbf{Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse: \emph{Demo: Composing,
Reproducing, and Sharing Simulations}}~\cite{Ganguly:2016ws}.
The authors demonstrate their approach of sharing,
reproducing, and composing simulations toward accelerating
research productivity while also improving accountability and
credibility. Specifically, they have developed a case study in
which they compose and share access control simulations in the
form of shareable data store units for cloud systems. They share, compose
and repeat simulations  through a collaborative repository
(OCCAM) and suggest a general simulation framework (SST) that can
accelerate the efforts for the whole community.

\item \textbf{Sameer Shende and Allen Malony: \emph{Demo: Using TAU for
Performance Evaluation of Scientific Software}}~\cite{Shende:2016ws}.
This demo presents the  TAU Performance System for performance evaluation of Scientific
Software written in C++, C, and Fortran. The system addresses performance
technology problems  at three levels: instrumentation,
measurement, and analysis. The TAU framework supports the
configuration and integration of these layers. The authors advocate that effective
exploration of performance will necessarily
require prudent selection from the range of alternative methods TAU provides
to  assemble meaningful performance experiments.

\item \textbf{Janos Sallai, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
\emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield
Atom-Typing for Molecular Simulation}}~\cite{Sallai:2016ws}.
The authors present an essential and error-prone task in the area of Molecular
Dynamics simulations: the efficient selection of correct parameter values for
specific molecules, especially forcefields. They present a framework that
aims to solve this data management issue, proposing a common format for
forcefields that is self-documenting with machine readable,
declarative usage rules. They investigate processes and tools
that are commonly used today in software development (e.g.,
unit testing, verification and validation, continuous integration,
and version control) and reason that they are, with proper infrastructure
support, applicable to forcefield development, as well. The
paper describes how such an infrastructure can tackle managing and evolving
forcefields by the molecular dynamics community, and
proposes a way to encourage and incentivize involvement
by the stakeholders.

@@

    \item \textbf{Sandra Gesing, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
    Suresh Marru, Marlon Pierce, Nancy Wilkins-Diehr, and Michael Zentner:
    \emph{The Science Gateways Community Institute - Supporting Communities to
    Achieve Sustainability for Their Science Gateways}}~\cite{Gesing:2016ws}.
    In this talk, Gesing introduced the concept of science gateways: end-to-end software
    solutions for a community that hide underlying complexity. She also posed the need
    for gateways, which include the increased complexity of research and associated
    hardware\slash software, a greater need for reproducibility and openness, and the
    opportunity to integrate research with teaching for better workforce development.
    In fact, gateway users surpassed login-based users in XSEDE systems in 2013,
    and 1.4 times as many users accessed XSEDE via gateways compared with
    login in March 2016. The Science Gateways Community Institute offers support
    related to gateways in five areas: Incubator, Extended Developer Support, Scientific
    Software Collaborative, Community Engagement and Exchange, and Workforce
    Development.

@@

% NSF grant for WSSSPE4
This material is based upon work supported by the National Science Foundation (ACI-1648293),
by the Gordon and Betty Moore Foundation (GBMF\#5620),
and by the Alfred P.~Sloan Foundation (G-2016-7214).

>>>>

%
For talks with multiple authors, an \textsuperscript{\textasteriskcentered} indicates the author who presented the talk.
%
%\todo{group by type - idea, position, experience, demo}

The following four papers belong to the first area and tackle diverse aspects from incentives for quality software to advocating a professional society for research software to roles and degrees for research software engineers.
\begin{itemize}
\item \textbf{Michael Heroux: \emph{Idea paper: Sustainable \& Productive:
Improving Incentives for Quality Software}}~\cite{Heroux:2016ws}.
The author presents that improving quality of scientific software is a measure
to address the increased complexity in both modeling and software. To support
quality, it is essential to provide Computational Science and Engineering teams
with publication venues, funding, and  professional recognition. Broad
improvements in sustainability and productivity can be only achieved when
publishers, funding agencies and employers raise their expectations for
software quality. Additionally, software community leaders are in the unique
position to positively impact software quality by working on establishing
incentives that will spur creative and novel approaches to improve developer
productivity  and software sustainability.

@@

\item \textbf{Olivier Philippe\textsuperscript{\textasteriskcentered}, Simon Hettrick, and Neil Chue Hong:
\emph{Experience Paper: Preliminary analysis of a survey of UK Research
Software Engineers}}~\cite{Philippe:2016ws}.
This paper presents results from a survey conducted on the new role of research
software engineers in academia. The survey provides demographic information
about the education, field, gender, job satisfaction, and career plans of the people
in this community. The outcome is that members of the community  were found to be
highly educated, derive mainly from the hard sciences, and are predominantly male.
Respondents reported satisfaction in their jobs, but indicated that career progression
is both difficult and opaque. The authors urge a continued discussion about the experience
of research software engineers and recommend further investigation into this
important community.

@@

The second main area concerned with characteristics and needs of research software includes the following five papers.

\item \textbf{Anshu Dubey\textsuperscript{\textasteriskcentered} and Katherine Riley: \emph{Experience Paper: Software
Engineering and Community Codes Track in ATPESC}}~\cite{Dubey1:2016ws}.
The Argonne Training Program in Extreme Scale
Computing (ATPESC) was started by Argonne National Laboratory with the objective
of expanding the ranks of better prepared users of high-performance computing (HPC) machines.
One of the unique aspects of the program was inclusion of a
software engineering and community codes track. The inclusion
was motivated by the observation that the projects with a good
software process were better able to meet their scientific goals.
The paper presents the experience in running the software
track, the motivation, reception, and evolution of the track over the years.
The authors recommend the inclusion of similar tracks in other HPC-oriented training programs.

@@

\item \textbf{Anshu Dubey\textsuperscript{\textasteriskcentered} and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
of Software for Scientific Simulation}}~\cite{Dubey2:2016ws}.
The authors discuss the specifics of the lifecycle of scientific software.
While the software lifecycle is a well-researched topic, one class of projects,
research software, has received relatively
little attention from lifecycle researchers. This
paper formalizes a lifecycle model
for end-to-end scientific application software, featuring loose
coupling between submodels for development of infrastructure
and scientific capability. The authors also invite input from stakeholders
to converge on a model that captures the complexity of this
development processes and provides needed lifecycle guidance to
the scientific software community.

\item \textbf{Francisco Queiroz\textsuperscript{\textasteriskcentered} and Rejane Spitz: \emph{Position Paper:
Collaborative Gamification Design for Scientific Software}}~\cite{Queiroz:2016ws}.
The paper focuses on usability design and presents that gamification, a design
trend for improving scientific software usability, for scientific software should
be facilitated by an open, collaborative design process supported by
conversational media. This approach is compatible with
qualities often attributed to computational science community
regarding openness and collaboration between members of varied
professional backgrounds. They exemplify the use of conversational media for
collaborative design and expect the synergy between collaborators to result in
better usability, greater user acceptance, and adequacy to requirements,
obtaining optimal design solutions in a sustainable way.

\item \textbf{Bruce Childers\textsuperscript{\textasteriskcentered}, Jack Davidson, Wayne Graves, Bernard Rous, and
David Wilkinson: \emph{Position Paper: Active Curation of Artifacts is Changing
the Way Digital Libraries will Operate}}~\cite{Childers:2016ws}.
The paper presents the implications of ``active curation systems'' that provide
executable access to
artifacts and experiments behind published results and enables
their reuse. These systems allow changing and repeating experiments to understand
how an innovation behaves in conditions beyond the ones described in a paper and
address reproducibility challenges. Active curation systems
also  enable accountability and accelerate research progress  by
giving access to complete experimental details. As these systems
take  hold,  it  is  important  to  understand  their  capabilities  and
how digital libraries should be integrated with them. The authors describe a
study underway to explore how best
to do this integration.

Three contributions present work on the third main area and elucidate about lessons learned from or visions for use cases of scientific software and communities working with and\slash or on a software package.

\item \textbf{Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse\textsuperscript{\textasteriskcentered}: \emph{Demo: Composing,
Reproducing, and Sharing Simulations}}~\cite{Ganguly:2016ws}.
The authors demonstrate their approach of sharing,
reproducing, and composing simulations toward accelerating
research productivity while also improving accountability and
credibility. Specifically, they have developed a case study in
which they compose and share access control simulations in the
form of shareable data store units for cloud systems. They share, compose
and repeat simulations  through a collaborative repository
(OCCAM) and suggest a general simulation framework (SST) that can
accelerate the efforts for the whole community.

\item \textbf{Sameer Shende\textsuperscript{\textasteriskcentered} and Allen Malony: \emph{Demo: Using TAU for
Performance Evaluation of Scientific Software}}~\cite{Shende:2016ws}.
This demo presents the  TAU Performance System for performance evaluation of Scientific
Software written in C++, C, and Fortran. The system addresses performance
technology problems  at three levels: instrumentation,
measurement, and analysis. The TAU framework supports the
configuration and integration of these layers. The authors advocate that effective
exploration of performance will necessarily
require prudent selection from the range of alternative methods TAU provides
to  assemble meaningful performance experiments.

\item \textbf{Janos Sallai\textsuperscript{\textasteriskcentered}, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
\emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield
Atom-Typing for Molecular Simulation}}~\cite{Sallai:2016ws}.
The authors present an essential and error-prone task in the area of Molecular
Dynamics simulations: the efficient selection of correct parameter values for
specific molecules, especially forcefields. They present a framework that
aims to solve this data management issue, proposing a common format for
forcefields that is self-documenting with machine readable,
declarative usage rules. They investigate processes and tools
that are commonly used today in software development (e.g.,
unit testing, verification and validation, continuous integration,
and version control) and reason that they are, with proper infrastructure
support, applicable to forcefield development, as well. The
paper describes how such an infrastructure can tackle managing and evolving
forcefields by the molecular dynamics community, and
proposes a way to encourage and incentivize involvement
by the stakeholders.

@@

    \item \textbf{Sandra Gesing\textsuperscript{\textasteriskcentered}, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
    Suresh Marru, Marlon Pierce, Nancy Wilkins-Diehr, and Michael Zentner:
    \emph{The Science Gateways Community Institute - Supporting Communities to
    Achieve Sustainability for Their Science Gateways}}~\cite{Gesing:2016ws}.
    In this talk, Gesing introduced the concept of science gateways: end-to-end software
    solutions for a community that hide underlying complexity. She also posed the need
    for gateways, which include the increased complexity of research and associated
    hardware\slash software, a greater need for reproducibility and openness, and the
    opportunity to integrate research with teaching for better workforce development.
    In fact, gateway users surpassed login-based users in XSEDE systems in 2013,
    and 1.4 times as many users accessed XSEDE via gateways compared with
    login in March 2016. The Science Gateways Community Institute offers support
    related to gateways in five areas: Incubator, Extended Developer Support, Scientific
    Software Collaborative, Community Engagement and Exchange, and Workforce
    Development.

@@

% NSF grant for WSSSPE4
This material is based upon work supported by the National Science Foundation (ACI-1648293, ACI-1547611),
by the Gordon and Betty Moore Foundation (GBMF\#5620),
and by the Alfred P.~Sloan Foundation (G-2016-7214).
